---
title: "biostat-200c-HW1-AC"
author: "Andrew Chuang"
format:
  html:
    theme: cosmo
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    cache: false    
    echo: true
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
editor: visual
---

## BIOSTAT 200C HW1 - Andrew Chuang

Session information for reproducibility purposes.
```{r}
sessionInfo()
```

### Libraries

```{r}
library(tidyverse)
library(faraway)
```
### Q1: Review of linear models

An initial data analysis that explores the numerical and graphical characteristics of the data.
        
An initial data analysis that explores the numerical and graphical 
characteristics of the data. I use the `summary` function in order to take a
first look at the basic numerical characteristics of the dataset.
```{r}
summary(swiss)
```

Variable selection to choose the best model.

An exploration of transformations to improve the fit of the model.

Diagnostics to check the assumptions of your model.

Some predictions of future observations for interesting values of the predictors.

An interpretation of the meaning of the model by writing a scientific abstract. (<150 words)

    BACKGROUND: brief intro of the study background, what are the existing findings

    OBJECTIVE: state the overall purpose of your research, e.g., what kind of knowledge gap you are trying to fill in

    METHODS: study design (how these data were collected), outcome definitions, statistical procedures used

    RESULTS: summary of major findings to address the question raised in objective

    CONCLUSIONS:
    
    